A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-05-04 submission/manuscript.rmd
Fix typo
changes based on previous reviewer comments

Understanding how a treatment’s effect varies across patients—a concept
described as heterogeneity of treatment effects (HTE)—has been central to the
agenda for both personalized (or precision) medicine and comparative
effectiveness research. More formally, HTE has been defined as the non-random
variability in the direction or magnitude of a treatment effect, in which the
effect is measured using clinical outcomes [@KRAVITZ2004]. Usually, analyses
focus on the relative scale, where treatment effects are assessed one at a time
in patient subgroups defined from single covariates, an approach that suffers
from low power and multiplicity issues [@Yusuf1991; @FernandezyGarcia2010].
However, even with well-established constant relative effects, treatment benefit
(or harm) may vary substantially on the absolute scale.

More recently, “predictive” HTE analyses have been described (and contrasted
with “one-variable-at-a-time” subgroup analysis) as approaches that provide
predictions of potential outcomes in a particular patient with one intervention
versus an alternative, taking into account multiple relevant patient
characteristics [@Varadhan2013]. One promising approach is “risk modeling”, in
which treatment effects are estimated in strata of predicted risk [@Kent2010;
@Kent2018]. Such a risk-based approach first stratifies patients according to
baseline risk predictions, using either an existing or an internally developed
risk prediction model [@Kent2010]. Then, relative and absolute treatment
effects are estimated within risk strata.

While these approaches have generally been recommended for application to
clinical trials, observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, and include heterogeneous populations. However, unlike trials,
treatment effects are subject to confounding while the unique structure of
different databases calls for database-specific analysis plans that are often
not easily transportable.

@@

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data. We implemented the
framework using existing OHDSI methods for use in the OMOP-CDM, including the
patient-level prediction framework and the population-level effect estimation
framework based on new-user cohort design [@Reps2018; @Ryan2013] . As a
proof-of-concept we analyzed heterogeneity of the effects of first-line
hypertension treatment: we compared the effect of angiotensin converting enzyme
(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A simple overview of the procedure
can be seen in Figure \@ref(fig:figure1).

@@

@@

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arms in order to avoid inducing spurious interactions
[@Burke2014; @vanKlaveren2019]. To do this, we develop the patient-level
prediction model in the propensity score-matched subset of the population (1:1),
where treatment assignment is well-balanced. The propensity scores are based on
LASSO logistic regression for modeling the association between treatment
assignment and all available demographics, drug exposures, diagnoses,
measurements and medical procedures. Finally, we need to define the time horizon
within which we aim to make predictions and we also need to select the
machine-learning algorithm we want to use to generate patient-level predictions.
Currently, the available options are regularized logistic regression, random
forest, gradient boosting machines, decision tree, naive Bayes, K-nearest
neighbors, neural network and deep learning (convolutional neural networks,
recurrent neural network and deep nets).

@@

We use the patient-level prediction model to divide the target population into a
set of equally-sized risk strata, typically 4 risk quarters. Then, we estimate
propensity scores within risk strata. These propensity scores are used when
estimating treatment effects, either by matching of patients from different
treatment cohorts, by stratification of patients into groups with similar
propensity scores, or by weighing patients’ contribution to the estimation
process. Within risk strata we estimate treatment effect both on the relative
and the absolute scale. It is important to evaluate treatment effects in both
scales, as effect cannot remain constant on both the relative and the absolute
scale at the same time, assuming a non-zero treatment effect. Any appropriate
method for the evaluation of relative and absolute treatment effects can be
considered, as long as the this is done consistently in all risk strata.

@@

The number of patients and person years by treatment arm along with the number
of outcomes should be reported. A performance overview of the derived prediction
models, including discrimination and calibration both in the propensity score
matched subset, the entire population and separately for treated and comparator
patients should also be reported. This is quite relevant as the performance of
the prediction models is directly related to our ability to single out patient
subgroups where treatment may be highly beneficial or unsafe. Kent et al
[@Kent2016] demonstrated that the event rate and the discriminative ability of
the prediction model can predict very well the distribution of predicted risk.
Lower event rate and higher c-statistic (given good calibration) result in high
risk heterogeneity, thus making estimated average treatment effects
uninformative. In this case, risk stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

In every risk stratum propensity score distributions by treatment group and
covariate balance plots should be made available along with event rates, hazard
ratios and absolute risk differences. Finally, the results should be generated
with a standardized format to allow for combination of results from multiple
data sources. This also enables the development of scalable demonstration tools
such as R-shiny applications, interactive tables and figures.

@@

We demonstrate the framework with the following research aim: "to compare the
effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients
with established hypertension ($D$) with respect to 9 outcomes
($O_1,\dots,O_9$)". The cohorts are:

@@

To obtain a target cohort for developing patient-level predictions we first
merged the ACE-inhibitors cohort with the beta blockers cohort. We then matched
patients in the ACE-inhibitor cohort to patients in the beta blockers cohort on
the propensity score. We focused on the efficacy outcomes (acute MI,
hospitalization with heart failure and hemorrhagic or ischemic stroke) for risk
stratification of the patient population. In each database, for each main
outcome, we developed a different prediction model. We chose a time horizon of 2
years after inclusion into the target cohort. We developed the prediction models
using LASSO logistic regression with 3-fold cross validation for hyper-parameter
selection.

@@

We used patient-level predictions to stratify the patient population into 4 risk
quarters. We estimated relative and absolute treatment effects for all 9
outcomes of interest. In order to estimate risk quarter-specific treatment
effects we first estimated propensity scores within risk quarters. We then used
the propensity scores to stratify patients into 5 strata.

@@

Here we present the results of the analysis in the CCAE database with
stratification based on risk predictions of acute MI. Results of analyses in the
other databases and with other risk stratifications are included in the
supplementary material.

@@

Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
ratios decreased) with increasing acute MI risk, resulting in more pronounced
absolute risk differences (ARD) with increasing acute MI risk (Figure
\@ref(fig:figure2)).

@@

Patients in the low risk quarter did not receive absolute benefit (-0.03%)
while absolute risk was 0.54% lower (95% confidence interval 0.36%—0.71%) for
patients in the high-risk quarter. In contrast, the absolute and relative
effects of ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are
approximately constant or even slightly decreasing with increasing acute MI risk
(Figure \@ref(fig:figure3) and \@ref(fig:figure4)). Similar results were observed
in the other two databases (see supplementary material).

@@

This example nicely illustrates heterogeneity of absolute treatment effects,
i.e. differences in absolute benefits and harms of ACE-inhibitors vs beta
blockers for patients with different baseline risk. The results suggest that
treatment with ACE-inhibitors, compared to treatment with beta blockers, may be
focused on the higher risk patients, in whom the benefits outweigh the harms.
However, treatment with beta blockers may be a viable option in lower risk
patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This
is in accordance with earlier findings that beta blockers should also be
considered as first-line treatment for younger hypertensive patients
[@Cruickshank2007; @Khan2006]. More thorough evaluation of these results is
required in future research, however.

@@

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are often adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale assessment of
treatment effect heterogeneity. Multiple outcomes can be evaluated in patient
subgroups of similar baseline outcome risk. Different outcome risk
stratification schemes can also be considered. The standardized nature of the
framework enables transportability to multiple databases, provided that they are
mapped to the OMOP-CDM.

Recently, guidelines on the application of risk modeling approaches for the
assessment of heterogeneity of treatment effect in RCT settings have been
proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
guidelines to the observational setting while also providing a toolset for its
implementation. Several considerations need to be made. First, estimates may be
biased due to the observational nature of the data. We attempt to account for
potential confounding by estimating propensity scores within strata of predicted
risk. These scores are estimated using regularized logistic regression on a
large set of pre-defined covariates. This specific approach gave accurate
results in extensive simulation studies [@Tian2018]. However, such approaches
do not account for unobserved confounding [@Liu2013]. Several sensitivity
analyses have been proposed in the literature for measuring the robustness of
results in the presence of unobserved confounding. Another approach is to
calibrate estimates and confidence intervals based on a large set of negative
controls [@Schuemie2014; @Schuemie2571]. Negative controls are
treatment-outcome pairs for which a null effect has been established. Estimating
these effects within available data provides an approximation of the null
distribution that can be used to empirically recalibrate effect estimates.
Future work may extend our framework with this type of analyses.

Our method provides a risk-stratified assessment of treatment effect
heterogeneity. However, even though stratification can provide a useful overview
for clinical interpretation, it is not appropriate to guide clinical practice,
where decisions need to be made at the individual level [@Kent2019].
Presentation of treatment effects as a continuous function of risk would be more
helpful, but is methodologically challenging. Future research is necessary for
the development of methods for continuous risk-based assessment of HTE.

@@

Our contribution is a translation of the PATH statement principles to the OHDSI
methods library [@Kent2019; @PathEnE]. Our method encourages open science as
it requires clear definition of the research questions translated into clear and
reproducible cohort definitions that can easily be shared among researchers. Our
R-package provides a standardized stepwise procedure for the assessment of HTE.
This enables source code to be easily shared and evaluated, allowing for
reproducible research. Researchers with access to different databases mapped to
OMOP-CDM can also very easily extend their overall analyses with risk-based
assessment of treatment effect heterogeneity. This enables collaboration among
multiple sites with access to different patient populations. We propose that the
framework is implemented any time treatment effect estimation in
high-dimensional observational data is undertaken.

>>>>

Understanding how a treatment’s effect varies across patients---a concept
described as heterogeneity of treatment effects (HTE)---has been central to the
agenda for both personalized (or precision) medicine and comparative
effectiveness research. More formally, HTE has been defined as the non-random
variability in the direction or magnitude of a treatment effect, in which the
effect is measured using clinical outcomes [@KRAVITZ2004]. Usually, analyses
focus on the relative scale, where treatment effects are assessed one at a time
in patient subgroups defined from single covariates, an approach that suffers
from low power and multiplicity issues [@Yusuf1991; @FernandezyGarcia2010].
However, even with well-established constant relative effects, treatment benefit
(or harm) may vary substantially on the absolute scale.

More recently, “predictive” HTE analyses have been described as approaches that
provide predictions of potential outcomes in a particular patient with one
intervention versus an alternative, taking into account multiple relevant
patient characteristics [@Varadhan2013]. We focus on “risk modeling”
approaches where patients are divided into risk strata using either an existing
or an internally developed risk prediction model. Risk-stratum-specific
estimates provide an overview of the evolution of treatment effects with
increasing risk both on the relative and the absolute scale. Recently,
systematic guidance on the application of such methods has been developed
[@Kent2010; @Kent2018].

While these approaches were developed for application in randomized controlled
trials (RCTs), observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, including heterogeneous populations. However, unlike RCTs,
treatment effects are subject to confounding while the unique structure of
different databases calls for database-specific analysis plans that are often
not easily transportable.

@@

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data, which extends the existing
guidelines of the RCT setting. We implemented the framework using existing OHDSI
methods for use in the OMOP-CDM, including the patient-level prediction
framework and the population-level effect estimation framework based on new-user
cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
heterogeneity of the effects of first-line hypertension treatment: we compared
the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
on 9 outcomes across three different US claims databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A overview of the procedure
can be seen in Figure \@ref(fig:figure1).

@@

Note "cohort" refers to a set of persons who satisfy one or more criteria for a
duration of time. The term can be used interchangeably with the term phenotype.

@@

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. To avoid deferentially fitting the prediction model to
patients across treatment arms, thus introducing spurious interactions with
treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
prediction model in the propensity score-matched (1:1) subset of the population.
Finally, we need to define the time horizon within which we aim to make
predictions and we also need to select the machine-learning algorithm we want to
use to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

@@

The aim of this step is the estimation of treatment effects (both on the
relative and the absolute scale) within risk strata---typically 4 risk
quarters---defined using the prediction model of step 3. Any appropriate method
for the evaluation of relative and absolute treatment effects can be considered,
as long as the this is done consistently in all risk strata. Common approaches
are odds ratios or hazard ratios for the relative scale and differences in
proportions or differences in Kaplan-Meier estimates on the absolute scale,
depending on the problem at hand. We estimate propensity scores within risk
strata which we then use to match patients from different treatment cohorts
or stratify them into groups with similar propensity scores or to weigh
each patient's contribution to the estimation process [@Austin2011].

@@

In the presence of a positive treatment effect and a well-discriminating
prediction model we expect an increasing pattern of the differences in the
absolute scale, even if the relative effects remain constant on the relative
scale across risk strata. Results should be presented side-by-side on a
forest-plot-like format, so that the evolution of treatment effects across risk
strata is visible both on the relative and the absolute scale.

A performance overview of the derived prediction models, including
discrimination and calibration both in the propensity score matched subset, the
entire population and separately for treated and comparator patients should also
be reported. This is quite relevant as the performance of the prediction models
is directly related to our ability to single out patient subgroups where
treatment may be highly beneficial or unsafe. Kent et al [@Kent2016]
demonstrated that the event rate and the discriminative ability of the
prediction model can predict very well the distribution of predicted risk. Lower
event rate and higher c-statistic (given good calibration) result in high risk
heterogeneity, thus making estimated average treatment effects uninformative. In
this case, risk stratified analysis of HTE can be more effective in singling out
patient subgroups that stand to benefit (or be harmed) most by treatment in
question.

In each risk stratum we need to evaluate if adequate covariate balance was
achieved accounting for measured confounding. If that is not the case,
interpretation of the results may be problematic. Common approaches include
evaluation of the overlap of propensity score distributions and calculation of
standardized covariate differences before and after propensity score adjustment.

@@

We consider the following research aim: "compare the effect of ACE-inhibitors
($T$) to the effect of beta blockers ($C$) in patients with established
hypertension ($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts
are:

@@

We developed separate prediction models for all efficacy outcomes (acute MI,
hospitalization with heart failure and hemorrhagic or ischemic stroke) in each
database. More specifically, we first estimated propensity scores using LASSO
logistic regression and a large set of baseline covariates including
demographics, drug exposures, diagnoses, measurements and medical devices. The
prediction models were estimated on the propensity score matched (1:1) subset of
the population, using caliper of 0.2 and after excluding patients having the
outcomes any time prior to treatment initiation. We chose a time horizon of 2
years after inclusion into the target cohort. For this demonstration, we
developed the prediction models using LASSO logistic regression with 3-fold
cross validation for hyper-parameter selection. We chose a time horizon of 2
years after inclusion into the target cohort. For this demonstration, we
developed the prediction models using LASSO logistic regression with 3-fold
cross validation for hyper-parameter selection.

@@

We used patient-level predictions to stratify the patient population into 4 risk
quarters. Within risk strata, relative effects were estimated using Cox
regression and absolute effects were estimated from the Kaplan-Meier estimate
differences at 2 years after treatment initiation. To adjust for observed
confounding within risk strata, we estimated propensity scores using the same
approach as in the development of prediction models. We used the estimated
propensity scores to stratify patients into 5 strata, within each risk quarter.

@@

We present the results of the analysis in the CCAE database with
stratification based on risk predictions of acute MI. Results of analyses in the
other databases and with other risk stratifications are included in the
supplementary material.

@@

Relative treatment effects of ACE-inhibitors vs beta blockers increased in favor
of ACE-inhibitors (hazard ratios decreased) with increasing acute MI risk,
resulting in more pronounced absolute risk differences (ARD) with increasing
acute MI risk (Figure \@ref(fig:figure2)).

@@

In general, with increasing acute MI risk we observed an increasing pattern in
terms of absolute benefit for the main outcomes, while the absolute harms did
not increase for the safety outcomes. More specifically, patients in the low
risk quarter did not receive absolute benefit (-0.03%) while absolute risk
was 0.54% lower (95% confidence interval 0.36%—0.71%) for patients in the
highest risk quarter. In contrast, the absolute and relative effects of
ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are slightly
decreasing with increasing acute MI risk (Figure \@ref(fig:figure3) and
\@ref(fig:figure4)). Similar results were observed in the other two databases
(see supplementary material).

@@

These results suggest that treatment with ACE-inhibitors, compared to treatment
with beta blockers, may be focused on the higher risk patients, in whom the
benefits outweigh the harms, while beta blockers may be a viable option in lower
risk patients, in whom the benefit-harm tradeoff is more favorable. This is in
accordance with earlier findings that beta blockers should also be considered as
first-line treatment for younger hypertensive patients [@Cruickshank2007;
@Khan2006]. More thorough evaluation of these results is required in future
research, however.

@@

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are often adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
in patient subgroups of similar baseline outcome risk. Different outcome risk
stratification schemes can also be considered. The standardized nature of the
framework enables transportability to multiple databases, provided that they are
mapped to the OMOP-CDM.

Recently, guidelines on the application of risk modeling approaches for the
assessment of heterogeneity of treatment effect in RCT settings have been
proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
guidelines to the observational setting while also providing a toolset for its
implementation. Several considerations need to be made. First, estimates may be
biased due to the observational nature of the data. We attempt to account for
potential confounding by estimating propensity scores within strata of predicted
risk. These scores are estimated using regularized logistic regression on a
large set of pre-defined covariates. However, such approaches do not account for
unobserved confounding [@Liu2013]. Several sensitivity analyses have been
proposed in the literature for measuring the robustness of results in the
presence of unobserved confounding. Another approach is to calibrate estimates
and confidence intervals based on a large set of negative controls
[@Schuemie2014; @Schuemie2571]. Negative controls are treatment-outcome pairs
for which a null effect has been established. Estimating these effects within
available data provides an approximation of the null distribution that can be
used to empirically recalibrate effect estimates. Future work may extend our
framework with this type of analyses.

Our method provides a risk-stratified assessment of treatment effect
heterogeneity. However, even though stratification can provide a useful overview
for clinical interpretation, these results cannot be applied to individuals in a straightforward manner, as we are still estimating subgroup effects [@Kent2019].
Presentation of treatment effects as a continuous function of risk would be more
helpful, but is methodologically challenging. Future research is necessary for
the development of methods for continuous risk-based assessment of HTE.

@@

Our contribution is a translation of the PATH statement principles to the OHDSI
methods library [@Kent2019; @PathEnE]. Our method encourages open science as
it requires accurate definition of the research questions translated into clear
and reproducible cohort definitions that can easily be shared among researchers.
Our R-package provides a standardized stepwise procedure for the assessment of
HTE. This enables source code to be easily shared and evaluated, allowing for
reproducible research. Researchers with access to different databases mapped to
OMOP-CDM can also very easily extend their overall analyses with risk-based
assessment of treatment effect heterogeneity. This enables collaboration among
multiple sites with access to different patient populations. We propose that the
framework is implemented any time treatment effect estimation in
high-dimensional observational data is undertaken.
